---
title: "Is There a Linear Association Between Potential Risk of Lead Exposure and Poverty Levels?"
author: "Megan Joseph"
thanks: "Project repository available at: [https://github.com/meganajoseph/math261a_paper1](https://github.com/meganajoseph/math261a_paper1)."
date: today
date-format: long
abstract: "Many Americans live below the poverty line. As such, one might be interested in the indicators of poverty. Using data from the Calenviroscreen 4.0 tool, I show how the potential risk of lead exposure in homes may impact the poverty levels in a census tract. I fit a simple linear regression model with the percent of the population living two times below the federal poverty line as the response and the potential risk of lead exposure to children in low income homes as the predictor. The output is the model Y = 9.29 + 0.45X where Y is the response and X is the predictor."
format: pdf
bibliography: references.bib
number-sections: true
---

```{r}
#| include: false
#| warning: false
#| message: false
# load libraries
library(tidyverse)
library(ggpubr)
library(patchwork)
```

# Introduction {#sec-intro}

Can the potential risk of lead exposure in homes indicate poverty levels in an area? Many Americans live below the federal poverty line, especially in California where the cost of living is high. Identifying indicators of poverty may bring us closer to ending it. It has already been shown that older housing, which often has lead-based paint, and poverty are associated with high blood lead levels [@Data_report].

\[include bg info on poverty and lead\]

\[address knowledge gap, analysis, analysis, findings\]

The remainder of the paper is structured as follows: @sec-data discusses the data, @sec-methods discusses the model, and @sec-results presents the results.

# Data {#sec-data}

```{r fig-calenviro-scatter}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot of the risk of lead exposure (x-axis) and percent of population in poverty (y-axis)."

library(readxl)
calenviroscreen <- read_excel("../data/calenviroscreen40resultsdatadictionaryf2021/calenviroscreen40resultsdatadictionary_F_2021.xlsx")

# dropping rows
calenviroscreen <- calenviroscreen[!with(calenviroscreen, is.na(Lead) & is.na(Poverty)),]

calenviroscreen$Lead = as.numeric(calenviroscreen$Lead)

ggplot(calenviroscreen, aes(x = Lead, y = Poverty)) +
  geom_point(color="deepskyblue3") +
  labs(x="Risk for lead exposure",
       y="Percent of population in poverty",
       title="Comparison of lead exposure and poverty for 2010 census\ntracts in California") +
  theme_minimal()
```

## Data Overview {#sec-data-overview}

The data is from CalEnviroScreen 4.0 which is a tool created by the Office of Environmental Health Hazard Assessment in the California Environmental Protection Agency used to help identify communities in California that are most affected by pollution [@Data_report]. The data used for the tool come from national and state sources and was last updated in 2021 [@Data].

Each of the `r nrow(calenviroscreen)` rows represents a census tract from 2010. The key variables are Lead, which quantifies the potential risk for lead exposure in children in low-income communities, and poverty, which represents the percent of population living below two times the federal poverty level [@Data_report]. Lead is transformed to a numeric variable because it is originally a character. Additionally, there are `r sum(is.na(calenviroscreen$Lead))` missing rows whicih will be dropped from the dataset. Many other variables are present in the dataset but will not be used for this paper. @fig-calenviro-scatter shows a scatter plot of Poverty on the y-axis and Lead on the x-axis. As the risk for lead exposure increases, the percent in poverty increases a bit as well.

## Poverty {#sec-poverty}

The poverty column was taken from the American Community Survey, an ongoing survey of a sample of US population by the US Census Bureau, from 2015-2019 [@Data_report]. While the survey uses the federal poverty level to determine if people live below the federal poverty line, this dataset uses twice the federal poverty line because of California's high cost of living [@Data_report]. Poverty is calculated by dividing the total population with a poverty status with the number of individuals below 200% of the poverty level for each census tract in California [@Data_report]. There are `r sum(is.na(calenviroscreen$Poverty))` missing rows which will be dropped from the data.

One issue with the dataset is that the American Community Survey data come from a sample of the population rather than the entire population, so the Poverty variable may not be fully representative of the California population. To increase reliability, census tracts with estimates that either had a relative standard error less then 50 or a standard error less than the mean standard error of all California census tract estimates for poverty were included and the rest received no score [@Data_report].

## Lead {#sec-lead}

Another potential issue is that there isn't direct information on exposures of lead, so the values used are based on pollution sources, releases, and environmental concentrations for potential exposures [@Data_report]. To calculate this value, they used the percentage of households in each census tract with a likelihood of lead-based paint hazards based on housing age and the percentage of low-income households with children [@Data_report]. The percentage of home with likelihood of lead-based paint hazards was calculated in a series of steps. First, California homes were grouped into categories based on age [@Data_report]. The, the number of housing units in each category was multiplpied by the reported percentage of homes with lead-based paint hazards [@Data_report]. Finally, the number of housing units in each category were added and divided by the total number of housing units in the census tract [@Data_report]. The percentage of low income households with children was calculated by estimating the number of households with incomes less than 80% of the county mean with one or more children under the age of six [@Data_report]. The final lead risk for housing score is the weighted sum of the percent of homes with likelihood of lead-based paint hazards percentile and the low-income household with children percentile: $0.6L + 0.4H = S$ where $L$ is the percentile for lead-based paint hazards, $H$ is the percentile of low-income household with children, and $S$ is the final score [@Data_report]. I will assume that the indicators created are indicative of the potential for lead exposure.

## Alternative Sources {#sec-alt-sources}

Alternative sources of data include finding data directly from the US Census so as to use data from the entire population rather than a sample. Using a dataset that had actual measures of lead exposure in housing would also lead to more reliable answers.

# Methods {#sec-methods}

To answer the research question, I will fit a simple linear regression model to the data with Poverty as the response and Lead as the predictor. Let $Y_i$ represent the percent of the population in the census tract living below two times the federal poverty level and $X_i$ represent the potential risk for lead exposure in children living in low-income communities with older housing. The model can we written as $$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \text{ for i = 1, ..., n}$$where $\beta_0$ represents the mean of the probability distribution of the percent of the population in poverty when risk for lead exposure is 0, $\beta_1$ represents the average increase in percent in poverty per unit increase in risk for lead exposure, and $\varepsilon_i$ represents random error with mean 0 and variance $\sigma^2$.

The assumptions for linear regression are:

1.  The variables used accurately reflect the quantities of interest, the model should include all predictors, and the model should be applicable for the research question.
2.  The sample data is representative of the population of interest.
3.  The mean function must be correct.
4.  Errors must be independent.
5.  The variance of the errors must be equal.
6.  Errors must be normal.

\[address assumptions\]

I use the `lm()` function from the R programming language [@R_language] to fit the linear regression model.

# Results {#sec-results}

```{r fig-model-fit}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot of the risk of lead exposure (x-axis) and percent of population in poverty (y-axis) with fitted linear regression model $Y_i = b_0 + b_0 X_i$."
model <- lm(calenviroscreen$Poverty ~ calenviroscreen$Lead)

ggplot(calenviroscreen, aes(x = Lead, y = Poverty)) +
  geom_point(color="deepskyblue3") +
  labs(x="Risk for lead exposure",
       y="Percent of population in poverty",
       title="Comparison of lead exposure and poverty for 2010 census\ntracts in California with fitted simple linear regression model") +
  geom_abline(aes(intercept=coefficients(model)[1], slope=coefficients(model)[2]), color="black") +
  theme_minimal()
```

The estimated slope parameter is $b_1=$ `r round(summary(model)$coefficients[2], 3)`. This means that for each unit increase in potential risk for lead exposure, the percent in poverty increases by `r round(summary(model)$coefficients[2], 3)` on average. The estimated intercept parameter is $b_0  =$ `r round(summary(model)$coefficients[1], 3)`. This means that when the potential risk for lead exposure is 0, the average percent in poverty is `r round(summary(model)$coefficients[1], 3)`.

\[add hypothesis test\]

\[add discussion\]

# References {#sec-references}
