---
title: "Is There a Linear Association Between Potential Risk of Lead Exposure and Poverty Levels?"
author: "Megan Joseph"
thanks: "Project repository available at: [https://github.com/meganajoseph/math261a_paper1](https://github.com/meganajoseph/math261a_paper1)."
date: today
date-format: long
abstract: "Many Americans live below the poverty line. As such, one might be interested in the indicators of poverty. Using data from the Calenviroscreen 4.0 tool, I investigate how the potential risk of lead exposure to children in low-income homes may impact the poverty levels in a California census tract. I fit a simple linear regression model with the percent of the population living two times below the federal poverty line as the response and the potential risk of lead exposure to children in low income homes as the predictor. A hypothesis test showed that there is a linear association, but it may be more impactful to do multiple regression because of the nuances surrounding poverty."
format: pdf
bibliography: references.bib
number-sections: true
---

```{r}
#| include: false
#| warning: false
#| message: false
# load libraries
library(tidyverse)
library(ggpubr)
library(patchwork)
library(ggplot2)
library(ggtext)
library(tidyr)
```

# Introduction {#sec-intro}

In 2023, around 6.4 million people in California were living below the California Poverty Measure poverty line (\$43,990 annually for two adults and two children) [@PPIC]. Identifying indicators of poverty may bring us closer to ending it. One potential indicator is lead exposure, although poverty is a very complex issue and is a culmination of a variety of factors. Children are more susceptible to the health effects that come from lead exposure due to how they are exposed, their developing brains, and how the lead is absorbed [@Data_report]. Many California houses were built before the lead-based paint ban in 1978 and still remains a a major source of lead exposure to young children [@Data_report]. Additionally, around 25% of children in California live in poverty. It has already been shown that older housing, which often has lead-based paint, and poverty are associated with high blood lead levels [@Data_report]. Based on this, I aim to answer the research question: is there a linear association between the potential risk for lead exposure in children in low-income communities and the percent of population living below two times the federal poverty level.

As a student pursuing their Masters in Statistics, there are many things I still don't know. I am not very well-versed in the social sciences which may impact the results of this study. Additionally, I am still learning about linear regression as a whole which may also impact the study.

In this study, I use data from CalEnviroScreen 4.0, specifically the Lead and Poverty columns, in an attempt to answer the research question. Then I fit a simple linear regression model with Poverty as the response and Lead as the predictor. Lastly, I perform a hypothesis test to determine if there really is a linear association between the two.

The results of the hypothesis test indicate that there is a linear association. This is important because it suggests that the potential risk for lead exposure can be used to measure poverty. Despite this, a multiple linear regression model may be a better fit for this study.

The remainder of the paper is structured as follows: @sec-data discusses the data, @sec-methods discusses the model, and @sec-results presents the results and their implications.

# Data {#sec-data}

```{r}
#| include: false
#| warning: false
#| message: false

library(readxl)
calenviroscreen <- read_excel("C:/Users/megan/OneDrive/Documents/math_261a/math261a_paper1/data/calenviroscreen40resultsdatadictionaryf2021/calenviroscreen40resultsdatadictionary_F_2021.xlsx")
```

## Data Overview {#sec-data-overview}

The data is from CalEnviroScreen 4.0 which is a tool created by the Office of Environmental Health Hazard Assessment in the California Environmental Protection Agency used to help identify communities in California that are most affected by pollution [@Data_report]. The data used for the tool come from national and state sources and was last updated in 2021 [@Data].

Each of the `r nrow(calenviroscreen)` rows represents a census tract from 2010. A census tract is made up of the smallest geographic unit for population data: census blocks [@Data_report]. They are typically areas bounded by streets, streams, and other features [@Rossiter_2021]. The key variables are Lead, which quantifies the potential risk for lead exposure in children in low-income communities, and poverty, which represents the percent of the population living below two times the federal poverty level [@Data_report]. Lead is transformed to a numeric variable because it is originally a character. Additionally, Lead ranges from 0 - 100 due to how it was created. See @sec-lead for more details. Many other variables are present in the dataset but will not be used for this paper. @fig-calenviro-scatter shows a scatter plot of Poverty on the y-axis and Lead on the x-axis. As the risk for lead exposure increases, the percent in poverty increases a bit as well.

Since this data is at the census tract level rather than at the household level, this analysis cannot conclude anything on the household scale, only at the census tract level. This analysis is not equivalent to an analysis at an individual household level. While there may be similar trends across both, the conclusions will differ. This is because we lose some variation at the household level when data is aggregated in this way. Since data is only available at the census tract level, the analysis will proceed at that scale knowing that analysis cannot be made on the household level.

```{r fig-calenviro-scatter}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot of the risk of lead exposure (x-axis) and percent of population in poverty (y-axis)."

# dropping rows
calenviroscreen$Lead <- as.numeric(calenviroscreen$Lead)
calenviroscreen <- calenviroscreen %>% drop_na(Lead, Poverty)

ggplot(calenviroscreen, aes(x = Lead, y = Poverty)) +
  geom_point(color="pink2") +
  labs(x="Risk for lead exposure",
       y="Percent of population in poverty",
       title="Comparison of lead exposure and poverty for 2010 census\ntracts in California") +
  theme_minimal()
```

## Poverty {#sec-poverty}

The poverty column was taken from the American Community Survey (ACS), an ongoing survey of a sample of the US population by the US Census Bureau, from 2015-2019 [@Data_report]. While the survey uses the federal poverty level to determine if people live below the federal poverty line, this dataset uses twice the federal poverty line because of California's high cost of living [@Data_report]. As a result, Poverty is calculated by dividing the total population with a poverty status with the number of individuals below 200% of the poverty level for each census tract in California rather than 100% [@Data_report].

One issue with the dataset is that the ACS data come from a sample of the population rather than the entire population, so the Poverty variable may not be fully representative of the California population. To increase reliability, census tracts with estimates that either had a relative standard error less than 50 or a standard error less than the mean standard error of all California census tract estimates for poverty were included and the rest received no score [@Data_report].

## Lead {#sec-lead}

Another potential issue is that there isn't direct information on exposures of lead, so the values used are based on pollution sources, releases, and environmental concentrations for potential exposures [@Data_report]. To calculate this value, they used the percentage of households in each census tract with a likelihood of lead-based paint hazards based on housing age and the percentage of low-income households with children [@Data_report]. The percentage of homes with a likelihood of lead-based paint hazards was calculated in a series of steps. First, California homes were grouped into categories based on age [@Data_report]. Then, the number of housing units in each category was multiplied by the reported percentage of homes with lead-based paint hazards [@Data_report]. Finally, the number of housing units in each category were added and divided by the total number of housing units in the census tract [@Data_report]. The percentage of low income households with children was calculated by estimating the number of households with incomes less than 80% of the county mean with one or more children under the age of six [@Data_report]. The final lead risk for housing score is the weighted sum of the percent of homes with likelihood of lead-based paint hazards percentile and the low-income household with children percentile: $0.6L + 0.4H = S$ where $L$ is the percentile for lead-based paint hazards, $H$ is the percentile of low-income households with children, and $S$ is the final score [@Data_report]. This creates a lowest value of 0 when both $L$ and $H$ are 0 and a highest value of 100 when both $L$ and $H$ are 100. I will assume that the indicators created are indicative of the potential for lead exposure.

## Alternative Sources {#sec-alt-sources}

Alternative sources of data include finding data directly from the US Census so as to use data from the entire population rather than a sample. Additionally, finding data on the household level would allow conclusions per household rather than per census tract. Using a dataset that had actual measures of lead exposure in housing would also lead to more reliable answers.

# Methods {#sec-methods}

To answer the research question, I will fit a simple linear regression model to the data with Poverty as the response and Lead as the predictor. Let $Y_i$ represent the percent of the population in the census tract living below two times the federal poverty level and $X_i$ represent the potential risk for lead exposure in children living in low-income communities with older housing. The model can we written as $$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \text{ for i = 1, ..., n}$$where $\beta_0$ represents the mean of the probability distribution of the percent of the population in poverty when risk for lead exposure is 0, $\beta_1$ represents the average change in percent in poverty per unit increase in risk for lead exposure, and $\varepsilon_i$ represents independent random error assumed to have mean 0 and variance $\sigma^2$.

```{r}
#| echo: false
#| warning: false
#| message: false
model <- lm(calenviroscreen$Poverty ~ calenviroscreen$Lead)
```

The assumptions for linear regression are:

1.  The variables used accurately reflect the quantities of interest, the model should include all predictors, and the model should be applicable for the research question.
2.  The sample data is representative of the population of interest.
3.  The mean function must be correct.
4.  Errors must be independent.
5.  The variance of the errors must be equal.
6.  Errors must be normal.

## Addressing Assumptions

1.  The variables used here are Poverty and Lead from the dataset. They reflect the quantities of interest: the percent living under two times the federal poverty line and the potential risk of lead exposure. The model includes Lead as a predictor.

2.  The population of interest is a California census tract. The sample data comes from each census tract in California which is representative of our population of interest. Certain areas, though, may not be accounted for in the Lead column due to ACS taking data from only a sample of the US population.

3.  There is a slight departure from linearity as seen in @fig-assumptions (a) where the residuals are not equally dispersed around the line $y=0$. This means that both the parameter estimates and error variance will be biased [@KNNL]. A curvilinear line may be better suited, but I will still fit a simple linear regression.

4.  Knowing information about one census tract will tell us information on the surrounding tracts because they do not contain huge geographic areas. Since their demographics are similar, their scores will be similar as well. As a result, the errors are not independent. This means that the estimators are unbiased but have biased variance [@KNNL].

5.  @fig-assumptions (a) shows that the variance of the errors may not be equal. It shows a slight cone shape, especially below the $y=0$ line. Nonconstancy of error variance is very common, but it does lead to less efficient estimates and invalid error variance estimates [@KNNL].

6.  A Q-Q plot of the residual percentiles against the standard normal percentiles is the best way to check for normality because we can see direct departures from the normal distribution. @fig-assumptions (b) shows that the errors appear to be close to normal with upward deviations on each end. I will assume that the errors are normal.

```{r fig-assumptions, fig.width=8, fig.height=4}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Plot a shows unequal variance in error terms and plot b shows deviations from normality on both ends of the residual percentile."

par(mfrow=c(1, 2))
var_errors <- ggplot(model, aes(x=.fitted, y=.resid)) +
  geom_point(color="pink2") +
  labs(x="Fitted values",
       y="Residuals",
       title="Checking constancy of error variance by\nplotting residuals against fitted values (a)") +
  geom_hline(aes(yintercept=0), color="pink4") +
  theme_minimal() +
  theme(plot.title=element_text(size=12))

normal_errors <- ggplot(model, aes(sample=.resid)) +
  stat_qq(color="pink2") +
  stat_qq_line(color="pink4") +
  labs(x="Standard normal percentile",
       y="Residuals percentile",
       title="Checking normality of errors with a Q-Q plot\nof residual percentile over standard normal\npercentile (b)") +
  theme_minimal() +
  theme(plot.title=element_text(size=12))

var_errors + normal_errors
```

I use the `lm()` function from the R programming language [@R_language] to fit the linear regression model.

# Results {#sec-results}

The estimated slope parameter is $b_1 \approx$ `r round(summary(model)$coefficients[2], 3)`. This means that for each unit increase in potential risk for lead exposure, the percent in poverty increases by `r round(summary(model)$coefficients[2], 3)` on average. The estimated intercept parameter is $b_0  \approx$ `r round(summary(model)$coefficients[1], 3)`. Since the dataset includes a potential risk for lead exposure at 0, the average percent in poverty at that value is `r round(summary(model)$coefficients[1], 3)`.

The coefficient of determination ($R^2$) is `r round(summary(model)$r.squared, 3)`. This means that `r 100 * round(summary(model)$r.squared, 3)`% of the variation in percent of households living two times below the federal poverty level is explained by the potential risk of lead exposure. This is not very high which is most likely due to poverty being a much more complex issue that cannot be explained by potential lead exposure alone. Other possible implications of poverty include, but are not limited to, race, education, neighborhood, and parent income levels. @fig-model-fit plots the regression line on top of the data. Many of the points are far above and far below the line which only heightens the fact that potential risk for lead exposure is not the only explanatory variable for poverty.

```{r fig-model-fit}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot of the risk of lead exposure (x-axis) and percent of population in poverty (y-axis) with fitted simple linear regression model $Y_i = b_0 + b_0 X_i$."

ggplot(calenviroscreen, aes(x = Lead, y = Poverty)) +
  geom_point(color="pink2") +
  labs(x="Risk for lead exposure",
       y="Percent of population in poverty",
       title="Comparison of lead exposure and poverty for 2010 census\ntracts in California with fitted simple linear regression model") +
  geom_abline(aes(intercept=coefficients(model)[1], slope=coefficients(model)[2]), color="pink4") +
  theme_minimal()
```

## Hypothesis Test

A two-sided hypothesis test of the form

$$H_0: \beta_1 = 0$$ $$H_a: \beta_1 \neq 0$$

will be used to test if there is a linear association between Poverty and Lead. If we conclude the null hypothesis, then there is no linear association [@KNNL].

To do this, we assume the normal error regression model [@KNNL]. Earlier, I found that the errors are approximately normal, but the variance does not appear to be constant. Since this does not make the regression invalid, I will proceed as if it were constant.

I will control the risk of a Type I error at $\alpha = 0.05$. With this, I will make a 95% confidence interval. If 0 is contained in the interval, then I will conclude $H_0$. Otherwise, I will conclude $H_a$ [@KNNL].

```{r}
#| echo: false
#| warning: false
#| message: false

mse <- sum(model$residuals**2) / (nrow(calenviroscreen) - 2)
s_squared <- mse / sum((calenviroscreen$Lead - mean(calenviroscreen$Lead))**2)
s <- sqrt(s_squared)
t <- qt(p=0.975, df=(nrow(calenviroscreen) - 2))

lower_bound <- model$coefficients[2] - t*s
upper_bound <- model$coefficients[2] + t*s
```

The 95% confidence limits for $\beta_1$ are $b_1 \pm t(1-\alpha/2;n-2)s\{b_1\}$ where $t(1 - \alpha/2; n-2)$ is the $\alpha/2$ 100 percentile of the t distribution with $n-2$ degrees of freedom and $s\{b_1\}$ is the point estimator of the standard deviation [@KNNL]. Plugging in our numbers, the 95% confidence limits for $\beta_1$ are `r round(model$coefficients[2], 3)` $\pm t(0.975;$ `r nrow(calenviroscreen) - 2` ) $\cdot$ `r round(s, 3)`

After calculating, `r round(lower_bound, 3)` $\leq \beta_1 \leq$ `r round(upper_bound, 3)`. Since zero is not contained in our interval, we can reject the null hypothesis. Therefore, there is a linear association between Poverty and Lead.

## Discussion {#sec-discussion}

The aim of this paper was to answer the following research question: Is there a linear association between potential risk of lead exposure and poverty levels? I investigated this with data from CalEnviroScreen 4.0. Then, I created a simple linear regression model using Lead as the predictor and Poverty as the response. The final model is $Y =$ `r round(model$coefficients[1], 3)` $+$ `r round(model$coefficients[2], 3)` $\cdot X$. The model explained `r 100 * round(summary(model)$r.squared, 3)`% of the variation in percent of households living two times below the federal poverty level is explained by the potential risk of lead exposure. Lastly, I performed a hypothesis test and created a 95% confidence interval in order to determine if there truly is a linear association. The resulting confidence interval is `r round(lower_bound, 3)` $\leq \beta_1 \leq$ `r round(upper_bound, 3)` which does not include zero showing that there is a linear association.

Since my findings show a linear association, the potential risk for lead exposure in children in low-income communities leads some indication of poverty, though not all of it.

A potential weakness of this study is that Lead may be correlated with Poverty since it is based on low-income housing. Additionally, an exponential model may be a better fit than a linear one based on the scatter plot, which also shows high variance.

Potential extensions include creating a transformed multiple regression model with a few predictors, possibly socioeconomic ones as well. Using data for lead exposure that does not take into account low-income communities is a potential improvement. This study can also be extended to the entire United States population rather than just California.

# References {#sec-references}
